---
title: 'Where to Hold a Demo?'
subtitle: 'seventh *Prague R Meetup*'
author: '[Jindra Lacko](mailto:jindra@jla-data.net)'
date: '<br>15.6.2018 @ KPMG'
output:
  xaringan::moon_reader:
    css: ["default", "JLA-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(sf)
library(tmap)
library(RCzechia)
library(tidyverse)
```
# Agenda

## Spatial data - what makes them special? 

## R packages for GIS work

## Example - calculating meeting place for a demonstration
---
# Spatial Data

### The Spatial Part
- point = coordinates X and Y
- line = multiple points with beginning and end
- polygon = multiple points joined in order
- raster (special format) = grid with start and cell size

### The Data Part
- information on the spatial part

### Example
- a tree grows on X and Y coordinates, it is a [spruce](https://en.wikipedia.org/wiki/Spruce) about 25 years old and it is infected by the [bark beetle](https://en.wikipedia.org/wiki/European_spruce_bark_beetle)  
- coordinates XYZ demark a polygon with LAU1 code CZ0100, it is called [Prague](https://en.wikipedia.org/wiki/Prague) and it has 1 294 513 inhabitants


---
# Coordinate Reference Systems

### Angular (Geographic)
- position of a point on globe in **angular** units (latitude, longitude)
- universally valid (can be used for any part of the globe)
- example: [EPSG:4321](https://epsg.io/4326) - WGS84
- used in navigation and by Google Maps, GPS (geocashing)

### Planar (Projected)
- position of a point in 2D space in **distance** units (easting, northing)
- locally valid (Earth is not flat; compromises are made in projection)
- example: [EPSG:5513](https://epsg.io/5513-1623) - ing. Křovák / S-JTSK
- used in geodetics, requirement of certain calculations

---
# package `sf`

Implementation of the [Simple Features](https://en.wikipedia.org/wiki/Simple_Features) ISO standard of spatial data in R.

### The `sf` class
- extends the `data.frame` class = allowing both "regular" data columns and a special column *geometry* with spatial attributes in a single R object
- all functions and techniques for `data.frame` still apply - e.g. `dplyr::inner_join()` and `dplyr::filter()`
- can be saved into database (I suggest [PostgreSQL](https://www.postgresql.org/)) and treated as tabular data

### Functions for class `sf`
- `st_transform()` - switch between coordinate reference systems
- `st_buffer()` - returns a buffer around a point or a polygon
- `st_centroid()` - returns center of a polygon as a point
- `st_intersect()` - returns intersection of two spatial objects
- and many others ... :)

---
# package `tmap`

A tool for map creation - accepts data in the `sf` format.
```{r tmap, eval = T, echo = F, cache = T, message = F, out.width = '100%', fig.align="center", fig.asp = 0.55, dpi = 150}

# load spatial data included in the tmap package
data("World", "metro")

# calculate annual growth rate
metro$growth <- (metro$pop2020 - metro$pop2010) / (metro$pop2010 * 10) * 100

# plot
tm_shape(World) +
  tm_polygons("income_grp", palette = "-Blues", 
    title = "Income class", contrast = 0.7, border.col = "gray30", id = "name") +
    tm_text("iso_a3", size = "AREA", col = "gray30", root = 3) +
tm_shape(metro) +
  tm_bubbles("pop2010", col = "growth", border.col = "black", 
    border.alpha = 0.5,
    breaks = c(-Inf, 0, 2, 4, 6, Inf) ,
    palette = "-RdYlGn",
    title.size = "Metro population (2010)", 
    title.col = "Annual growth rate (%)",
    id = "name",
    popup.vars = c("pop2010", "pop2020", "growth")) + 
tm_format_World() + 
tm_style_gray()


```

---
# package `RCzechia`

Spatial objects of the Czech Republic (okresy, kraje, obce etc.) in the `sf` format.
```{r rczechia, eval = T, echo = T, cache = T, out.width = '80%', fig.align="center", fig.asp = 0.55, dpi = 150}
plot(RCzechia::okresy("low"), max.plot = 1)

```
---
# Planning a Demo

## **Input** - official election results by Czech municipalities

## **Output** - a municipality in the Czech republic with the most voters in its neighborhood

## **Parameter** - distance (radius of the "neighborhood")

---
# Plan of Action

### Acquire data
- election results: [CZSO.cz](https://volby.cz/opendata/prez2018/prez2018_opendata.htm)
- shapefile: `RCzechia::obce_polygony()` a `RCzechia::casti()`
- connecting the dots: `dplyr::inner_join()`

### Initial overview
- visualization: `tmap::tm_shape() + tmap::tm_fill()`

### Prepare function *Šelmostroj*
- brute force approach: neigborhood for every Czech municipality (6 387 in total)
- find the one with max total of votes in neigborhood
- report on the finding: print on console + draw a chart

### Apply function *Šelmostroj*
- iterate with different values of the *distance* parameter
---
# Relative Totals
Election result as *% of the vote* - the most common visualization...
```{r map-relative, eval = T, echo = F, cache = T, out.width = '100%', fig.align="center", fig.asp = 0.55, dpi = 150}

# načtení dat z jlacko/Zeman2018 na GitHubu
Volby2018 <- url("https://raw.githubusercontent.com/jlacko/Zeman2018/master/src/prezident.csv")
src <- read.csv2(Volby2018,
                 stringsAsFactors = F) 

# výsledky voleb
druheKolo <- src %>%
  filter(CHYBA == 0) %>% # bez chyb
  filter(KOLO == 2) %>% # pouze druhé kolo
  group_by(OBEC) %>%
  summarize(hlasu = sum(HLASY_07),  # kandidát č.7 = Miloš Zeman
            procent = sum(HLASY_07)/sum(PL_HL_CELK)) %>% # procento platných pro Zemana
  mutate(KOD = as.character(OBEC)) # kod obce v RCzechia je text :(

obce <- obce_polygony() %>% 
  select(KOD = KOD_OBEC,
         NAZEV = NAZ_OBEC)

casti <- casti() %>%
  select(KOD, NAZEV)

podklad <- obce %>% # všechny obce...
  rbind(casti) %>% # ...plus všechny části
  inner_join(druheKolo, by = c("KOD", "KOD")) %>%
    # z obcí a částí připojit ty s výsledkem
    # filtrační (inner) join odstraní obce bez výsledku (Praha etc. - má ho z částí)
  st_transform(crs = 5514) 
    # systém inž. Křováka 

st_agr(podklad) <- 'constant'
  # všechny hodnoty jsou konstanty

republika <- republika() %>% # obrysy republiky - pro hezčí grafy...
  st_transform(crs = 5514) 
    # systém inž. Křováka 

kraje <- kraje() %>% # obrysy krajů - pro hezčí grafy...
  st_transform(crs = 5514)
    # systém inž. Křováka 

velka_voda <- reky() %>% # více významné řeky
  filter(NAZEV %in% c('Vltava', 'Labe', 'Morava')) %>% 
  st_transform(crs = 5514)

mala_voda <- reky() %>% # méně významné řeky
  filter(NAZEV %in% c('Berounka', 'Sázava', 'Jizera', 'Otava', 'Ohře', 'Odra', 'Dyje', 'Svratka', 'Mže')) %>%
  st_transform(crs = 5514)


mapRelative <- tm_shape(podklad) + tm_fill(col = "procent", title = "Election result", n = 5) +
  tm_shape(kraje) + tm_borders("grey80") +
  tm_shape(republika) + tm_borders("grey35") +
  tm_shape(velka_voda) + tm_lines(col = 'steelblue', lwd = 1.5, alpha = 0.6) + 
  tm_shape(mala_voda) + tm_lines(col = 'steelblue', lwd = 1.25, alpha = 0.4) +
  tm_style_white(frame = F, 
                 legend.format = list(text.separator =  "-",
                                      fun = function(x) paste0(formatC(x * 100, digits = 0, 
                                                                       format = "f"), " %")),
                 legend.text.size = 0.8, 
                 legend.title.size = 1.3) +
  tm_legend(position = c("RIGHT", "top")) 

mapRelative

```

---
# Absolute Totals
Election result as *total number of votes* - because people matter...
```{r map-absolute, eval = T, echo = F, cache = T, out.width = '100%', fig.align="center", fig.asp = 0.55, dpi = 150}

mapAbsolute <- tm_shape(podklad) + tm_fill(col = "hlasu", title = "Election result", n = 5) +
  tm_shape(kraje) + tm_borders("grey80") +
  tm_shape(republika) + tm_borders("grey35") +
  tm_shape(velka_voda) + tm_lines(col = 'steelblue', lwd = 1.5, alpha = 0.6) + 
  tm_shape(mala_voda) + tm_lines(col = 'steelblue', lwd = 1.25, alpha = 0.4) +
  tm_style_white(frame = F, 
                 legend.format = list(text.separator =  "-",
                                      text.align = "center",
                                      fun = function(x) paste0(formatC(x / 1000, digits = 0, 
                                                                       format = "f"), " tis.")),
                 legend.text.size = 0.8, 
                 legend.title.size = 1.3) +
  tm_legend(position = c("RIGHT", "top"))

    #poznámka: tento mapový objekt ještě několikrát zrecykluji...

mapAbsolute
```
```{r housekeeping, eval = T, echo = F, warning = F, cache = T}
podklad <- podklad %>%  
  st_centroid() 
  # body místo polygonů - snazší výpočet vzdálenosti, a jednoznačná interpretace průsečíků 
```
---
# Šelmostroj function
```{r selmostroj, eval = T, echo = T, cache = T, message = F, warning = F, out.width = '100%', fig.align="center", fig.asp = 0.55, dpi = 150}
selmostroj <- function(vzdalenost) {

  for (i in 1:nrow(podklad)) { 
    buff <- st_buffer(podklad[i,], dist = vzdalenost) # buffer o průměru vzdalenost
    isect <- st_intersection(podklad, buff) # průsečík bufferu a pracovního seznamu
    podklad$suma_volicu[i] <- sum(isect$hlasu) # uložit součet hlasů v bufferu do podkladu
  }
  
  stred <- podklad[which.max(podklad$suma_volicu), ] # tady!

  writeLines(paste("The best place to hold the demo:", stred$NAZEV, 
            "\nVoters in", vzdalenost / 1000, "kilometer neighborhood:",
            formatC(stred$suma_volicu, format = "f", big.mark = " ", digits = 0)))
 
  mapAbsolute <- mapAbsolute + # recyklace dříve vytvořené mapy
    tm_shape(stred) + tm_dots(size = 1/5, col = "red", shape = 4) +
    tm_shape(st_buffer(stred, dist = vzdalenost)) + tm_borders(col = "red")

  print(mapAbsolute)
}
```
---
# 2 kilometer range
```{r spad-2, eval = T, echo = T, cache = T, message = F, warning = F, out.width = '85%', fig.align="center", fig.asp = 0.55, dpi = 150}
selmostroj(2 * 1000) # 2 kilometry
```
---
# 10 kilometer range
```{r spad-10, eval = T, echo = T, cache = T, message = F, warning = F, out.width = '85%', fig.align="center", fig.asp = 0.55, dpi = 150}
selmostroj(10 * 1000) # 10 kilometrů
```
---
# 30 kilometer range
```{r spad-30, eval = T, echo = T, cache = T, message = F, warning = F, out.width = '85%', fig.align="center", fig.asp = 0.55, dpi = 150}
selmostroj(30 * 1000) # 30 kilometrů
```
---
# 50 kilometer range
```{r spad-50, eval = T, echo = T, cache = T, message = F, warning = F, out.width = '85%', fig.align="center", fig.asp = 0.55, dpi = 150}
selmostroj(50 * 1000) # 50 kilometrů
```
---
# 150 kilometer range
```{r spad-150, eval = T, echo = T, cache = T, message = F, warning = F, out.width = '85%', fig.align="center", fig.asp = 0.55, dpi = 150}
selmostroj(150 * 1000) # 150 kilometrů
```
---
# Conclusion - R as a GIS tool 

## ✘ more code and less GUI
## ✔ reproducible and reusable
## ✔ lends itself to automation
## ✔ R ecosystem & community
## ☺ Open Source - for good or ill

---
class: inverse, middle, center

# Thank you for your attention

This presentation (including R Markdown source code) is on GitHub
[https://github.com/jlacko/r-meetup-demonstrace](https://github.com/jlacko/r-meetup-demonstrace)
